---
title: "250220_Calderon"
author: "Muhunden Jayakrishnan"
date: "2/20/2025"
output: html_document
---

Exploratory Deep Learning analyses of Calderon et al., 2023 Science 

Objective : 1) Manipulate scRNAseq data to generate inputs for DL modelling
            2) Compare NN architecture used in the manuscript against more complicated models

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(Seurat)
library(patchwork)
library(dplyr)
library(SeuratData)
```


## Load data 

Load data and generate folds for Deep Learning modelling.

For this purpose, we will use the pre-generated Seurat object which has equally sampled # of cells within each time window. Equal sampling prevents modelling biases towards few more-abundant classes.

```{r Data}

rna_subsampled <- readRDS("./SeuratObjects/RNA/14912_29824_subsampled_large_checkpoint_endpoint.Rds")

head(rna_subsampled@meta.data,n=10) ## Types of preexisting columns 

head(rna_subsampled@assays$RNA@counts,n=20)[,1:30] ##example RNA counts for some cells 

head(rna_subsampled@assays$RNA$data)  ## The data slot has depth normalized counts (as they are no longer integers!) 
```
Note that time column labels the original collection window -> We will use this information to generate the target variable (midpoint of each window) for modelling.

We will use depth normalized counts as features. Feature scaling will be performed before training. 


First verify if the RNA data is ordered correctly as the cell metadata labels.

```{r}
all(rna_subsampled@meta.data$cell == colnames(rna_subsampled@assays$RNA@data))

```

For modelling, we need to convert the Seurat object into a sparse-matrix containing relevant data.

```{r}
sparse_mat <- rna_subsampled@assays$RNA@data

sparse_mat_center <- rna_subsampled@meta.data %>% mutate(start=as.numeric(gsub('_.+$', '', gsub('hrs_', '', time))),
                                                          end=as.numeric(gsub('^.+_', '', gsub('hrs_', '', time))),
                                                          center=(start+end) / 2) %>% select(cell, center, time)


head(sparse_mat_center)
```

Feature selection : Remove low variance genes (preserving sparsity)

```{r}

sds_vals <- sparseMatrixStats::rowSds(sparse_mat)

head(sds_vals)

```

Also note that they mapped to an extended reference list which also included transposable elements and other non-gene annotations. Standard RNA-seq is not appropriate for these quantifications, so remove these and retain only well annotated genes.

```{r}

all_genes <- rownames(sparse_mat)

filtered_genes <- all_genes[which(sds_vals > 0 & !grepl('-transposable-element', all_genes) & !grepl('^FBti', all_genes))]

summary(filtered_genes) 

all(sparse_mat_center$cell == rownames(t(sparse_mat[filtered_genes,]))) ##sanity check 

```
We retain around 16k genes 

Now generate 11-kfold sets of cells (10 folds for train-test and 11th fold will be for validation).
Each fold will be stored as a .mtx sparse-matrix format which can be used in Python.
```{r}
k_folds <- 11 
sparse_mat_center$kfold <- ggplot2::cut_number(sample(nrow(sparse_mat_center)),
	k_folds, labels=1:k_folds)

fold_dir <- "./k_folds_data/"

if(!file.exists(fold_dir)){
  dir.create(fold_dir)
}

## loop through folds and write sparse object and save center windows

## transpose so features (genes) are columns and samples (cells) are rows
for (a_fold in 1:length(unique(sparse_mat_center$kfold))) { # a_fold <- 1
	print(paste0('working on fold: ', a_fold))
	fold_cells <- sparse_mat_center %>% filter(kfold == a_fold)
	Matrix::writeMM(obj = t(sparse_mat[filtered_genes, fold_cells$cell]),
		file=paste0(fold_dir,a_fold, "_fold_rna.mtx"))
	fold_cells %>% select(center, time, cell, kfold) %>%
	    write.table(paste0(fold_dir,a_fold, "_fold_rna_centers.tsv"),
			sep='\t', quote=F, col.names=F, row.names=F)
}

```
 Now process the main Seurat dataset object (containing all cells from all expts) in the same way to exclude unnecessary genes and extract 'centers' of time windows.
 
```{r}

temp <- readRDS(file = "SeuratObjects/RNA/main.Rds")

sparse_mat_full <- temp@assays$RNA@data
sparse_mat_full_center <- temp@meta.data %>% mutate(start=as.numeric(gsub('_.+$', '', gsub('hrs_', '', time))),
                                                    end=as.numeric(gsub('^.+_', '', gsub('hrs_', '', time))),
                                                    center=(start+end) / 2) %>% select(cell, center, time)

Matrix::writeMM(obj = t(sparse_mat_full[filtered_genes,]), file=paste0(fold_dir,"all_rna.mtx"))
sparse_mat_full_center %>% 
    write.table(paste0(fold_dir,"all_rna_centers.tsv"),
	sep='\t', quote=F, col.names=F, row.names=F)

```
 
